Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking

### 2023-08-01 manuscript/manuscript.qmd
revise based on comments from Jan

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we, in this
current study, considered as the gold standard for measuring sleep.
Furthermore, our secondary goal was to assess the developed models'
performance in evaluating important sleep quality parameters, including
sleep period time (SPT), total sleep time (TST), sleep efficiency (SE),
latency until persistent sleep (LPS), and wake after sleep onset (WASO).

@@

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
signal processing algorithms, employs three self-adhesive, disposable
sensors placed outside the hairline for reliable EEG signal acquisition.
The participants of the SCREENS study were instructed to attach the
device when they went to bed. The ZM uses two proprietary algorithms:
Z-ALG and Z-PLUS. The Z-ALG is utilized for accurate sleep detection,
showcasing its suitability for in-home monitoring[@kaplan2014], while
the Z-PLUS effectively differentiates sleep stages, as evidenced by its
alignment with expert evaluations using PSG data[@wang2015]. In the
current study, we treated all sleep stages as a single category
effectively deducing the output of the ZM to "awake" and "asleep" as the
ability to distinguish sleep stages are not a necessity to derive the
sleep quality parameters of interest and to simplify the learning
process of the models.

@@

In this study, we employed two distinct learning strategies to analyze
sleep patterns from thigh-mounted accelerometer data. The first strategy
employed a sequential methodology, using an ensemble of four pairs of
models, with each pair featuring the same algorithm. This sequential
strategy was designed to simplify the prediction task by decomposing it
into two stages: first predicting 'in-bed' time, then 'sleep' time. The
output from the first set of models, which predicted in-bed time, was
subjected to a 5-minute median filter to eliminate leading and trailing
in-bed time blips. This process enabled us to extract a single
continuous time interval that we identified as the sleep period time
window (SPT). The SPT, in turn, served as the input to the second models
in the sequence, enhancing their predictive accuracy for sleep time. We
applied this sequential strategy across all four algorithms, as detailed
below.

1.  Logistic Regression (LREG): Logistic regression served as a simple
    and fast baseline model. However, due to its linear nature, it may
    struggle with capturing complex relationships and non-linear
    patterns present in the accelerometer data.

2.  Decision Tree (TREE): Decision trees are capable of handling
    non-linear patterns and are easily interpretable. However, they are
    prone to overfitting, particularly when dealing with complex
    patterns that require simultaneous consideration of multiple
    features.

3.  Single-layer Feed-forward Neural Network (SNN): Single-layer
    feed-forward neural networks can effectively capture non-linear
    relationships, even with their relatively simple structure. However,
    they tend to be more challenging to interpret compared to simpler
    models. Additionally, careful tuning of the network's architecture
    and training process is required to mitigate the risk of
    overfitting.

4.  XGBoost (XGB): XGBoost is a powerful algorithm known for its ability
    to provide highly accurate predictions and handle complex,
    non-linear patterns in the data. It also incorporates built-in
    methods to prevent overfitting. However, training XGBoost models can
    be computationally intensive, and interpreting the predictions it
    generates can pose challenges.

In parallel, we also employed a multiclass approach using a
bidirectional Long Short-Term Memory (biLSTM)[@hochreiter1997] neural
network. This network, which was designed to predict three distinct
classes: 'out-of-bed-awake', 'in-bed-awake', and 'in-bed-asleep', was
configured with four layers and 128 hidden units per layer. This balance
between model complexity and training efficiency was intended to
facilitate learning of intricate patterns while ensuring feasible
training times. The bidirectional nature of the LSTM enhanced data
interpretation and reduced overfitting by doubling the hidden units at
each time step. The LSTM model used sequences of tensors as input, with
each sequence spanning 10 minutes and a step size of one. As
demonstrated by previous studies such as those by Sano et al.
[@sano2019] and Chen et al. [@chen2021], LSTM models have shown great
promise in sleep detection using accelerometer data, thanks to their
ability to capture complex temporal patterns.

@@

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately 50%
of the subjects. The splitting of the data was ensured to not have
samples from the same night simultaneously present in both sets. To
optimize hyperparameters, we performed a 10-fold Monte Carlo
cross-validation on a regular grid, comprising 20 different combinations
of hyperparameters. The F1 score served as the optimization metric. The
best-performing set of hyperparameters was then used to fit the models
to the full training dataset. This approach allowed us to maximize
performance by leveraging all available data. Moreover, after extracting
the in-bed time from the initial sequential models, the imbalance on the
resulting dataset could cause biases during model training, as models
may favor predicting the majority class. To rectify this, we employed
the Synthetic Minority Over-sampling Technique (SMOTE)[@chawla2002].
SMOTE generates new samples by interpolating random samples with their
nearest neighbors. We utilized the themis R package[@themis] to
implement SMOTE, resulting in a balanced distribution of training
samples across both classes.

@@

In our study, we utilized standard evaluation metrics to assess the
performance of each model on an epoch-to-epoch basis. These include
accuracy ($accuracy = \frac{TP+TN}{TP+TN+FP+FN}$), sensitivity
($sensitivity = \frac{TP}{TP+FN}$), specificity
($specificity = \frac{TN}{TN+FP}$), precision
($precision = \frac{TP}{TP+FP}$), negative predictive value (NPV,
$NPV = \frac{TN}{TN + FN}$), and F1 score
($F_1 = 2 * \frac{precision * sensitivity}{precision + sensitivity}$).

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs. out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs. awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using the
same metrics. To do this, we considered the multiclass output as to
binary classifications, where the first was out-of-bed vs the rest and
the second binary classification as in-bed-awake vs in-bed-asleep. To
further illustrate model performance, we provide confusion matrices for
the full dataset, encompassing both in-bed and out-of-bed data. These
matrices report relative counts, column percentages (the proportion of
the true class accurately predicted), and row percentages (the
proportion of predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research[@hjorth2012; @kushida2001].

To assess the performance of our models in deriving sleep quality
parameters, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Considering our dataset
contained multiple observations per subject, we integrated a bootstrap
procedure to address this extra source of variability. We calculated the
mean difference (bias) and defined the LOA as the mean difference plus
or minus 1.96 times the standard deviation of these differences. To
ensure our measurements were robust and accounted for intra-subject
variability, we estimated the 95% confidence intervals for both the bias
and the LOA using a bias-corrected and accelerated bootstrap method,
utilizing 10,000 bootstrap replicates. The sleep quality parameteres
included are defined as follews in accordance with the ZM definitions:

@@

As reported in @tbl-zm_overview the sleep quality parameters derived
from ZM predictions were modified by the implementation of 5-minute and
10-minute median filters. SPT were consistent across raw and filtered
datasets (mean: 9.2 ± 2.1 hours), corresponding to the length of the ZM
recording. TST and SE increased in the filtered data, implying the
filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. LPS
also elevated, suggesting the filter smooths out brief awakenings at
sleep onset, leading to a prolonged time to persistent sleep. A
significant change was seen in WASO, which dropped from 39.0 ± 33.6
minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in
the 5-minute and 10-minute filtered data, respectively. The number of
awakenings was also considerably reduced with the application of
filters. In the raw data, the average number of awakenings was 34.46 ±
11.33 per night, which reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the
5-minute and 10-minute filtered data sets respectively.

@@

The epoch-to-epoch evaluation of predicting in-bed time is outlined in
@tbl-in_bed_performance, and demonstrates practically equivalent
performance across all model types. The F1 score ranges from 94.4%
(Decision Tree) to 95.4% (XGBoost), while accuracy ranges from 95.3%
(Decision Tree) to 96.1% (XGBoost). Sensitivity, Precision, and
Specificity also demonstrate consistent results across the models. The
XGBoost model, despite recording the highest metrics with an F1 score of
95.4% and accuracy of 96.1%, outpaced the others only marginally.

@@

## Evaluation of Sleep Quality Parameters

@tbl-ba_cor presents a comparative analysis of the included models used
to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
using the 5-minute median filtered ZM predictions. To see the full table
including models developed from raw ZM predictions and 10-minute median
filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
decision tree model consistently underestimated SPT, TST, and SE, and
overestimated LPS and WASO in comparison to ZM. The logistic regression
model had similar trends, with more pronounced underestimation in TST
and overestimation in LPS. The feed-forward neural network also
exhibited similar bias as the decision tree and the logistic regression
models, but with a higher overestimation in WASO. On the other hand, the
XGBoost model showed least bias among all, especially in its 5-minute
median predictions. Considering LOA, the decision tree had higher
variability across different sleep quality parameters and filtering
techniques, particularly for LPS and WASO, which indicates lower
agreement with ZM. Other models had comparable LOA but with notable
exceptions. For example, TST LOA for the logistic regression model was
particularly wide in the 5-minute median predictions. Correlation-wise,
the pearson coefficient, revealed that the XGBoost model consistently
had the highest correlation with ZM across all sleep quality parameters
and filtering methods Notably, the XGBoost's 5-minute median predictions
showed the strongest correlation (0.66) for TST among all models and
filtering techniques.

@@

@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
on 5-minute median filtered ZM predictions, and the 5-minute
median-smoothed ZM-derived sleep quality parameters. The Bland-Altman
plot for the SPT and TST reveals a significant level of agreement with
the ZM, as evidenced by a bias close to zero. Interestingly, a portion
of the data points are located near the zero line indicating perfect
agreement. However, the presence of extreme outliers widens the limits
of agreement (LOA). The scatterplot for SPT also demonstrates a positive
trend, indicating a moderate linear correlation between the XGBoost
model and the ZM-derived sleep quality parameters. The bias and LOA for
TST are comparable to those observed for SPT, indicating a consistent
level of agreement between the two methods. The scatterplot for TST also
shows a slightly higher correlation, primarily driven by the absence of
extreme outliers.Furthermore, the remaining three sleep quality
parameters, SE, LPS, and WASO, exhibit heteroscedasticity in contrast to
SPT and TST. This outcome is expected as achieving 100% sleep efficiency
is relatively rare, resulting in less disagreement between the methods
as values approach the upper limit. However, as sleep efficiency
decreases, the potential for discrepancies and differing interpretations
between the methods increases, leading to greater heteroscedasticity. A
moderate positive linear correlation is observed between the XGBoost
model and ZM-derived sleep quality parameters for SE, however, a poor
correlation is observed for LPS and WASO, suggesting less agreement
between the two methods for these parameters. Similar plots for all
models are available in the **supplementary materials (75 plots. Jeg ved
ikke om nogen orker at bladre igennem dem).**

![Comparison of sleep quality parameters derived from the XGBoost model
trained on the 5-minute smoothed ZM predictions. The left column
displays Bland-Altman plots. Dashed lines represent the bias (the
average difference between the two measurements) and LOA, with the 95%
confidence intervals represented as the grayed areas. The right column
displays scatter plots of XGBoost-derived vs ZM-derived sleep quality
parameters. The dashed line represents the identity line, while the
full-drawn line represents the best linear fit. Pearson's correlations
are annotated in the upper left
corner.](visuals/median_5_xgboost_ba_cor.pdf){#fig-xgb_ba_cor}

@@

In an effort to mature the methods for estimating sleep from thigh-worn
accelerometers, we evaluated various models for predicting in-bed and
sleep time and their derived sleep quality parameters. Furthermore, we
trained and evaluated the models using raw and median-filtered gold
standard predictions from the ZM. In general, all sequential models
performed well at predicting in-bed time. More challenging was it to
distinguish wake from sleep on the extracted in-bed time, however, the
performance of the sequential models were enhanced by the application of
median filters. Moreover, even though the multiclass biLSTM showed good
performance across F1 score, precision and NPV, the derived sleep
quality parameters were not on par with the XGBoost model which
demonstrated the highest performance metrics across all evaluations,
including epoch-to-epoch prediction and sleep quality parameter
derivatives. Despite this, all sequential models showed low specificity
values, indicating difficulty in correctly classifying awake epochs
during time in bed. The application of 5-minute and 10-minute median
filters improved the performance metrics of all models, with the XGBoost
model consistently leading. The median filters increased total sleep
time and sleep efficiency, while reducing wake after sleep onset and the
number of awakenings. The XGBoost model also showed the least bias and
highest correlation with ZM across sleep quality parameters.

>>>>

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Dette er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we, in this
current study, considered as the gold standard for measuring sleep.
Furthermore, our secondary goal was to assess the developed models'
performance in evaluating important sleep quality metrics, including
sleep period time (SPT), total sleep time (TST), sleep efficiency (SE),
latency until persistent sleep (LPS), and wake after sleep onset (WASO).

@@

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
signal processing algorithms, employs three self-adhesive, disposable
sensors placed outside the hairline for reliable EEG signal acquisition.
The participants of the SCREENS study were instructed to attach the
device when they went to bed. The ZM uses two proprietary algorithms:
Z-ALG and Z-PLUS. The Z-ALG is utilized for accurate sleep detection,
showcasing its suitability for in-home monitoring[@kaplan2014], while
the Z-PLUS effectively differentiates sleep stages, as evidenced by its
alignment with expert evaluations using PSG data[@wang2015]. In the
current study, we treated all sleep stages as a single category
effectively deducing the output of the ZM to "awake" and "asleep" as the
ability to distinguish sleep stages are not a necessity to derive the
sleep quality metrics of interest and to simplify the learning process
of the models.

@@

We employed two different model strategies to assess sleep patterns from
thigh-mounted accelerometer data. The first model strategy was designed
as a sequence of two models to simplify the prediction task by
decomposing it into two stages: first predicting 'in-bed' time, then
'sleep' time. In the sequential model the output from the first set of
models, which predicted in-bed time, was subjected to a 5-minute median
filter to eliminate leading and trailing in-bed time blips. This process
enabled us to extract a single continuous time interval that we
identified as the sleep period time window (SPT). The SPT, in turn,
served as the input to the second models in the sequence, enhancing
their predictive accuracy for sleep time. We applied this sequential
strategy using the following four machine learning algorithms:

1.  Logistic Regression: Logistic regression served as a simple and fast
    baseline model. However, due to its linear nature, it may struggle
    with capturing complex relationships and non-linear patterns present
    in the accelerometer data.

2.  Decision Tree: Decision trees are capable of handling non-linear
    patterns and are easily interpretable. However, they are prone to
    overfitting, particularly when dealing with complex patterns that
    require simultaneous consideration of multiple features.

3.  Single-layer Feed-forward Neural Network: Single-layer feed-forward
    neural networks can effectively capture non-linear relationships,
    even with their relatively simple structure. However, they tend to
    be more challenging to interpret compared to simpler models.
    Additionally, careful tuning of the network's architecture and
    training process is required to mitigate the risk of overfitting.

4.  XGBoost: XGBoost is a powerful algorithm known for its ability to
    provide highly accurate predictions and handle complex, non-linear
    patterns in the data. It also incorporates built-in methods to
    prevent overfitting. However, training XGBoost models can be
    computationally intensive, and interpreting the predictions it
    generates can pose challenges.

In parallel, we also employed a multiclass algorithm as the second model
strategy using a bidirectional Long Short-Term Memory
(biLSTM)[@hochreiter1997] neural network which also incorporates
temporal aspects of the data. This network, which was designed to
predict three distinct classes: 'out-of-bed-awake', 'in-bed-awake', and
'in-bed-asleep', was configured with four layers and 128 hidden units
per layer. This balance between model complexity and training efficiency
was intended to facilitate learning of intricate patterns while ensuring
feasible training times. The bidirectional nature of the LSTM enhanced
data interpretation and reduced overfitting by doubling the hidden units
at each time step. The LSTM model used sequences of tensors as input,
with each sequence spanning 10 minutes and a step size of one. As
demonstrated by previous studies such as those by Sano et al.
[@sano2019] and Chen et al. [@chen2021], LSTM models have shown great
promise in sleep detection using accelerometer data, thanks to their
ability to capture complex temporal patterns.

@@

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately 50%
of the subjects. The splitting of the data was ensured to not have
samples from the same night simultaneously present in both sets. To
optimize hyperparameters, we performed a 10-fold Monte Carlo
cross-validation on a regular grid, i.e., for each hyperparameter, a
range of values at evenly-spaced intervals was selected. comprising 20
different combinations of hyperparameters. The F1 score served as the
optimization metric. The best-performing set of hyperparameters was then
used to fit the models to the full training dataset. This approach
allowed us to maximize performance by leveraging all available data. An
imbalance was observed with the in-bed time determined in the initial
step of the sequential model strategy which after extracting the in-bed
time from the initial sequential models, the imbalance on the resulting
dataset could cause biases during model training, as models may favor
predicting the majority class. To account for this imbalance, we
employed the Synthetic Minority Over-sampling Technique
(SMOTE)[@chawla2002]. SMOTE generates new samples by interpolating
random samples with their nearest neighbors. We utilized the themis R
package[@themis] to implement SMOTE, resulting in a balanced
distribution of training samples across both classes.

@@

In our study, we utilized standard evaluation metrics to assess the
performance of each model on an epoch-to-epoch basis. These include
$$accuracy = \frac{TP+TN}{TP+TN+FP+FN}$$
$$sensitivity = \frac{TP}{TP+FN}$$ $$specificity = \frac{TN}{TN+FP}$$
$$precision = \frac{TP}{TP+FP}$$ $$NPV = \frac{TN}{TN + FN}$$
$$F_1 = 2 \cdot \frac{precision \cdot sensitivity}{precision + sensitivity}$$

where NPV is negative predictive value, TP is true positives, FP is
false positives, TN is true negatives, and FN is false negatives.

In the context of our sequential model strategy, the initial models were
tasked with the binary classification of in-bed vs. out-of-bed. For this
task, we assessed performance using the F1-score, accuracy, sensitivity,
specificity, and precision metrics. The second models in our sequential
model strategy focused on the binary classification of asleep vs. awake.
For these models, we considered the same metrics, in addition to the
negative predictive rate. The class imbalance in this case led us to
compute the F1 score as an unweighted macro-average. Additionally, we
evaluated the multiclass classifier, biLSTM, using the same metrics. To
do this, we considered the multiclass output as to binary
classifications, where the first was out-of-bed vs the rest and the
second binary classification as in-bed-awake vs in-bed-asleep. To
further illustrate model performance, we provide confusion matrices for
the full dataset, encompassing both in-bed and out-of-bed data. These
matrices report relative counts, column percentages (the proportion of
the true class accurately predicted), and row percentages (the
proportion of predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research[@hjorth2012; @kushida2001].

To assess the performance of our models in deriving sleep quality
metrics, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Given the nature of our
dataset, which contains multiple observations per subject, we employed a
bootstrap procedure to account for this added variability. We first
calculated the mean difference (bias) and then defined the limits of
agreement (LOA) as the mean difference plus or minus 1.96 times the
standard deviation of these differences. Acknowledging the possibility
of non-normality and potential skewness in our data, we chose to apply a
bias-corrected and accelerated (BCa) bootstrap
method[@diciccio_bootstrap_1996]. This approach allowed us to better
address potential bias in our estimates and the inherent intra-subject
variability. Utilizing 10,000 bootstrap replicates, we estimated the 95%
confidence intervals for both the bias and the LOA, thus ensuring
robustness in our measurements. The sleep quality parameteres included
are defined as follows in accordance with the ZM definitions:

@@

As reported in @tbl-zm_overview the sleep quality metrics derived from
ZM predictions were modified by the implementation of 5-minute and
10-minute median filters. SPT were consistent across raw and filtered
datasets (mean: 9.2 ± 2.1 hours), corresponding to the length of the ZM
recording. TST and SE increased in the filtered data, implying the
filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. LPS
also increased, suggesting the filter removes brief awakenings at sleep
onset, leading to a prolonged time to persistent sleep. A change was
seen in WASO, which dropped from 39.0 ± 33.6 minutes in raw data to 30.6
± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and 10-minute
filtered data, respectively. The number of awakenings was also
considerably reduced with the application of filters. In the raw data,
the average number of awakenings was 34.46 ± 11.33 per night, which
reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
filtered data sets respectively.

@@

The epoch-to-epoch evaluation of predicting in-bed time is outlined in
@tbl-in_bed_performance, and demonstrates practically equivalent
performance across all model types. The F1 score ranges from 94.4%
(Decision Tree) to 95.4% (XGBoost), while accuracy ranges from 95.3%
(Decision Tree) to 96.1% (XGBoost). Sensitivity, Precision, and
Specificity also demonstrate consistent results across the different
models. The XGBoost model provide the best performance with an F1 score
of 95.4% and accuracy of 96.1%, although only outpacing the other models
marginally.

@@

## Evaluation of sleep quality metrics

@tbl-ba_cor presents a comparative analysis of the included models used
to predict various sleep quality metrics (SPT, TST, SE, LPS, WASO) using
the 5-minute median filtered ZM predictions. To see the full table
including models developed from raw ZM predictions and 10-minute median
filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
decision tree model consistently underestimated SPT, TST, and SE, and
overestimated LPS and WASO in comparison to ZM. The logistic regression
model had similar trends, with more pronounced underestimation in TST
and overestimation in LPS. The feed-forward neural network also
exhibited similar bias as the decision tree and the logistic regression
models, but with a higher overestimation in WASO. On the other hand, the
XGBoost model showed least bias among all, especially in its 5-minute
median predictions. Considering LOA, the decision tree had higher
variability across different sleep quality metrics and filtering
techniques, particularly for LPS and WASO, which indicates lower
agreement with ZM. Other models had comparable LOA but with notable
exceptions. For example, TST LOA for the logistic regression model was
particularly wide in the 5-minute median predictions. Correlation-wise,
the pearson coefficient, revealed that the XGBoost model consistently
had the highest correlation with ZM across all sleep quality metrics and
filtering methods Notably, the XGBoost's 5-minute median predictions
showed the strongest correlation (0.66) for TST among all models and
filtering techniques.

@@

@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
on 5-minute median filtered ZM predictions, and the 5-minute
median-smoothed ZM-derived sleep quality metrics. The Bland-Altman plot
for the SPT and TST reveals a good level of agreement with the ZM, as
evidenced by a bias close to zero. Interestingly, a portion of the data
points are located near the zero line indicating perfect agreement. The
scatterplot for SPT also demonstrates a positive trend, indicating a
moderate linear correlation between the XGBoost model and the ZM-derived
sleep quality metrics. The bias and LOA for TST are comparable to those
observed for SPT, indicating a consistent level of agreement between the
two methods. The scatterplot for TST also shows a slightly higher
correlation, primarily driven by the absence of extreme
outliers.Furthermore, the remaining three sleep quality metrics, SE,
LPS, and WASO, exhibit heteroscedasticity in contrast to SPT and TST.
This outcome is expected as achieving 100% sleep efficiency is
relatively rare, resulting in less disagreement between the methods as
values approach the upper limit. Conversely, as sleep efficiency
decreases, the potential for discrepancies and differing interpretations
between the methods increases, leading to greater heteroscedasticity. A
moderate positive linear correlation is observed between the XGBoost
model and ZM-derived sleep quality metrics for SE, however, a poor
correlation is observed for LPS and WASO. Similar plots for all models
are available in the **supplementary materials (75 plots. Jeg ved ikke
om nogen orker at bladre igennem dem).**

![Comparison of sleep quality metrics derived from the XGBoost model
trained on the 5-minute smoothed ZM predictions. The left column
displays Bland-Altman plots. Dashed lines represent the bias (the
average difference between the two measurements) and LOA, with the 95%
confidence intervals represented as the grayed areas. The right column
displays scatter plots of XGBoost-derived vs ZM-derived sleep quality
metrics. The dashed line represents the identity line, while the
full-drawn line represents the best linear fit. Pearson's correlations
are annotated in the upper left
corner.](visuals/median_5_xgboost_ba_cor.pdf){#fig-xgb_ba_cor}

@@

To select the most optimal method for estimating sleep from thigh-worn
accelerometers, we evaluated various models for predicting in-bed and
sleep time and their derived sleep quality metrics. We trained and
evaluated the models using raw and median-filtered gold standard sleeå
estimates from the ZM EEG-based sleep monitor. In general, all
sequential models performed well at predicting in-bed time. More
challenging was it to distinguish wake from sleep on the extracted
in-bed time, and the performance of the sequential models were enhanced
by the application of median filterings. Moreover, even though the
multiclass biLSTM showed good performance across F1 score, precision and
NPV, the derived sleep quality metrics were not on par with the XGBoost
model which demonstrated the highest performance metrics across all
evaluations, including epoch-to-epoch prediction and all sleep quality
metrics. Despite this, all sequential models showed low specificity
values, indicating difficulty in correctly classifying awake epochs
during time in bed. The application of 5-minute and 10-minute median
filters improved the performance metrics of all models. Median
filterings increase total sleep time and sleep efficiency, while
reducing wake after sleep onset and the number of awakenings. The
XGBoost model provides the smallest bias and highest correlation with
all ZM sleep quality metrics.

Limited research exists regarding the epoch-to-epoch effectiveness of
classifying in-bed time based on data from thigh-worn accelerometers.
Nevertheless, Carlson and colleagues provided compelling insights. They
demonstrated that a third-party algorithm, "ProcessingPal," and a
proprietary one, "CREA," achieved accuracies of 91% and 86%
respectively. These algorithms, evaluated against self-reported
measures[@carlson2021], produced F1 scores as high as 95% and 96%. These
figures are consistent with the performance of our sequential models,
which also achieved F1 scores and accuracy scores exceeding 95% in
identifying in-bed time. In our study, in-bed time is equated with SPT.
All models, with the exception of XGBoost, underestimated SPT. The
biLSTM model showed the greatest underestimation, with a bias of -36
minutes, reflecting trends observed in previous research. Winkler et al.
developed an algorithm that, despite a strong correlation (Pearson
correlation coefficient = .67) between their algorithmic results and
diary-recorded waking times, overestimated waking wear time by more than
30 minutes, resulting in an underestimation of in-bed
time[@winkler2016]. This trend was further confirmed when Inan-Eroglu et
al. examined Winkler et al.'s algorithm, revealing a underestimation of
9.8 minutes in bed time compared to self-reported
measures[@inan-eroglu2021]. In contrast, a study by van der Berg et al.
reported a slight underestimation of in-bed time. They employed a unique
approach with their algorithm, which relied on quantifying the number
and duration of sedentary periods to determine time in bed, and active
periods (standing or stepping) to identify wake times[@vanderberg2016].
Finally. it is important to note that high predictive performance in
determining in-bed time does not necessarily translate to accurate
predictions of broader sleep quality metrics. The crucial task of
detecting awake periods during in-bed time, a key factor in assessing
sleep quality, may not be effectively captured by in-bed time
predictions alone. Indeed, underestimating in-bed time could result in
overestimating waking time during in-bed time. Furthermore, the
distinction between actual sleep and time spent in bed, often overlooked
but vital in sleep research, is critical for a comprehensive
understanding of sleep quality.
